---
title: "Genetic exchange in Leishmania is facilitated by a natural antibody: Statistics"
author: "Johannes Doehl"
date: "2023-07-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)

# Libraries to be loaded

library(readxl) # Required to read in all worksheets in one go
library(xlsx) # required for loading and saving .xlsx files
library(rlist) # Required for list manipulation functions like list.append()
library(rcompanion) # Required for the pairwise comparison after mxn Fisher's Exact test [ pairwiseNominalIndependence() ]
library(rstatix) # Contains pipe friendly analysis functions
library(magrittr) # required for %$%
library(tidyverse) # multi-package that contains several required package inc. ggplot2, purrr, tipple, dplyr, etc.
```

## Functions defined by me

This chunk contains two functions that are not common to any packages and are thus defined by me for use in the script.

```{r Function-definition}
# Simultaneous data loading from multiple worksheet in same excel files
read_all <- function(filename, tibble = TRUE) {
    # If tibble is preferred over data.frame use "tibble = TRUE"
    sheets <- readxl::excel_sheets(filename)
    x <- lapply(sheets, function(X) readxl::read_excel(filename, sheet = X))
    if(!tibble) x <- lapply(x, as.data.frame)
    names(x) <- sheets
    x
}

# zipFastener for TWO dataframes of unequal length; by Mark Heckmann
zipFastener <- function(df1, df2, along=2)
{
  # parameter checking
  if(!is.element(along, c(1,2))){
    stop("along must be 1 or 2 for rows and columns respectively")
  }
  # if merged by using zip feeding along the columns, the
  # same no. of rows is required and vice versa
  if(along==1 & (ncol(df1)!= ncol(df2))) {
    stop ("the no. of columns has to be equal to merge them by zip feeding")
  }
  if(along==2 & (nrow(df1)!= nrow(df2))) {
    stop ("the no. of rows has to be equal to merge them by zip feeding")
  }
  
  # zip fastener preparations
  d1 <- dim(df1)[along]
  d2 <- dim(df2)[along]
  i1 <- 1:d1           # index vector 1
  i2 <- 1:d2 + d1      # index vector 2
  
  # set biggest dimension dMax
  if(d1==d2) {
    dMax <- d1
  } else if (d1 > d2) {
    length(i2) <- length(i1)    # make vectors same length, 
    dMax <- d1                  # fill blanks with NAs   
  } else  if(d1 < d2){
    length(i1) <- length(i2)    # make vectors same length,
    dMax <- d2                  # fill blanks with NAs   
  }
  
  # zip fastener operations
  index <- as.vector(matrix(c(i1, i2), ncol=dMax, byrow=T))
  index <- index[!is.na(index)]         # remove NAs
  
  if(along==1){
    colnames(df2) <- colnames(df1)   # keep 1st colnames                  
    res <- rbind(df1,df2)[ index, ]  # reorder data frame
  }
  if(along==2) res <- cbind(df1,df2)[ , index]           
  
  return(res)
}
```

## Variable information

This chunk contains variables that require adjustments to work with different experiments and data directories.

```{r Variable-information}
# File Paths
BiblioPath1 <- "C:/Users/doehljs/Box/Clump quantification/Dose response_Experiment/" # Root directory

Output <- "Data Analysis Output/" # Folder for data output

FilNam1 <- "Tiago/Data_Summary_In_Vivo_Clump_LLJB_Raw.xlsx" # Location of raw data file
```

## Data loading

Read in the raw data table from the excel workbook

```{r Data-loading}
# Loads all worksheets from xlsx
ClmpDat <- read_all(paste0(BiblioPath1, FilNam1))

# Vector of names of experiments (names from the xlsx file worksheets)
FigNam <- names(ClmpDat)
```

## Data reformating into contingency tables 

This chunk reformat the data from the excel spreadsheets into contingency tables by day of observation, where applicable 

```{r Contingency-tables}
options(scipen=5) # sets numbers of decimals permitted before scientific nomenclature takes over

# Data reformatting into contingency tables

# Creates a list to store the reformatted data into contingency table
ClmpTabList <- list()

# Summarizes categorical data into contingency tables
for (aa in 1:length(ClmpDat)){
  
  Days <- unique(ClmpDat[[aa]]$Day) # gets Days of measurement
  Outcome <- unique(ClmpDat[[aa]]$Result_1) # gets result descriptors
  ExpGrp <- unique(ClmpDat[[aa]]$Group) # Gets group names
  ListFrq <- list();
  
  # Gets the data summary tables
  for (bb in 1: length(Days)) {
    ByCnt <- ClmpDat[[aa]] %>%
      filter(Day == Days[bb]) %$%
      table(Result_1, Group)
    
    if ("Negative" %in% rownames(ByCnt) == FALSE) {
      RowAdd <- as.table(rep(0, times = ncol(ByCnt)))
      names(RowAdd) <- colnames(ByCnt)
      ByCnt <- rbind(RowAdd, ByCnt)
      rownames(ByCnt) <- Outcome
      ByCnt <- as.table(ByCnt)
      
    } else if ("Positive" %in% rownames(ByCnt) == FALSE) {
      RowAdd <- as.table(rep(0, times = ncol(ByCnt)))
      names(RowAdd) <- colnames(ByCnt)
      ByCnt <- rbind(ByCnt, RowAdd)
      rownames(ByCnt) <- Outcome
      ByCnt <- as.table(ByCnt)
    }
    
    # Creats percentage of count table by column
    ByPctCol <- as.data.frame.array(round(prop.table(ByCnt, margin = 2), 3) * 100)
    
    # Creats percentage of count table by row
    ByPctRow <- as.data.frame.array(round(prop.table(ByCnt, margin = 1), 3) * 100)
    
    ByCnt <- as.data.frame.array(ByCnt)
    
    ListFrq[[bb]] <- list(ByCnt, ByPctCol, ByPctRow)
    
    names(ListFrq[[bb]]) <- c("Count", "By_Column", "By_Row")
  }
  names(ListFrq) <- paste("Day", Days)
  
  ClmpTabList <- list.append(ClmpTabList, ListFrq)
}
names(ClmpTabList) <- FigNam

# Removes unwanted variables
rm(aa, bb, ByCnt, ByPctCol, ByPctRow, ListFrq)

gc() # Garbage collection
```

## Fisher's Exact test

This chunk performs the Fisher's exact test on the data

```{r Fishers-exact}
# Creates a list to store the test output data
ClmpFshList <- list()

# This loop executes the Fisher's exact test from stats with m x n tables and executes a pairwise comparison if indicated 
for (aa in 1:length(ClmpTabList)) {
  ListFT <- list()
  Days <- unique(names(ClmpTabList[[aa]]))
  
  for (bb in 1:length(Days)) {
    # Fisher's exact test for a 2 x 4 table of count data
    tryCatch({fisher.test(ClmpTabList[[aa]][[bb]][["Count"]], simulate.p.value = TRUE); EM <<- 0}, # In case one roq is all zeros
             error = function(e) {EM <<- 1});
    
    if (EM == 0) {
      # Fisher's Exact test for R x C table
      FT <- ClmpTabList[[aa]][[bb]][["Count"]] %>% 
        # { if ("Mice" %in% colnames(.) == TRUE) select(., -Mice) else . } %>% # In case of exclusion of the "Mice" data column 
        fisher_test(., simulate.p.value = TRUE) 
      
      if (FT$p < 0.05) {
        # Pairwsie comparaison as a post hoc test to Chi Square / Fisher's exact test with table exceeding 2 x 2 format
        pwcFT <- ClmpTabList[[aa]][[bb]][["Count"]] %>% 
          # { if ("Mice" %in% colnames(.) == TRUE) select(., -Mice) else . } %>% # In case of exclusion of the "Mice" data column 
          pairwiseNominalIndependence(.,
                                      compare = "column",
                                      fisher = TRUE, # Fisher's Exact test executed for the pair
                                      gtest = FALSE,
                                      chisq = FALSE,
                                      cramer = TRUE, # Cramer's V test determines the effect size between two measurements
                                      method = "fdr")
        
        # This rounds the values to 5 decimals
        pwcFT[, -which(names(pwcFT) %in% "Comparison")] <- round(pwcFT[, -which(names(pwcFT) %in% "Comparison")], digits = 5)
        
        # Creates a letter based comparison; same letter means no difference
        tryCatch({cldList(p.adj.Fisher ~ Comparison, data = pwcFT); EM <<- 0},
             error = function(e) {EM <<- 1});
        if (EM == 0) {
          pwcLet <- cldList(p.adj.Fisher ~ Comparison, data = pwcFT)
          
          # Compile data into a list
        ListFT[[bb]] <- list(FT, pwcFT, pwcLet)
        names(ListFT[[bb]]) <- c("Fisher's Exact", "Pairwise", "Letter Code")
  
        } else {
          # Compile data into a list
        ListFT[[bb]] <- list(FT, pwcFT)
        names(ListFT[[bb]]) <- c("Fisher's Exact", "Pairwise") 
        }
         
      } else {
        ListFT[[bb]] <- list(FT)
        names(ListFT[[bb]]) <- c("Fisher's Exact")
      }
      
    } else {
      FT <- list("p.value" = 1)
      ListFT[[bb]] <- list(FT)
        names(ListFT[[bb]]) <- c("Fisher's Exact")
    }
  }
  names(ListFT) <- Days
  
  ClmpFshList <- list.append(ClmpFshList, ListFT)
}
names(ClmpFshList) <- FigNam

rm(aa, bb, FT, pwcFT, pwcLet, ListFT, EM)

gc()
```

## Saving test outputs

This chunk saves the test's output to an excel file in a desired location. Note that locations need to be adjusted as desired.

```{r Saving-data}
# Will remove saved xlsx spreadsheets, if they already exist so the can be replaced as overwrite does not work here
if (file.exists(paste0(BiblioPath1, Output, "Stats/Statisitcs Table (Jul 23).xlsx")) == TRUE) { 
  
  sapply(paste0(BiblioPath1, Output, "Stats/Statisitcs Table (Jul 23).xlsx"), unlink);
}

# Create workbook to store data
saveWorkbook(createWorkbook(), paste0(BiblioPath1, Output, "Stats/Statisitcs Table (Jul 23).xlsx"))

# Load data from list into workbook
for (aa in 1:length(ClmpFshList)) {
  
  Days <- unique(names(ClmpTabList[[aa]]))
  
  for (bb in 1:length(Days)) {
    
    lapply(names(ClmpFshList[[aa]][[bb]]), function(x) write.xlsx(ClmpFshList[[aa]][[bb]][[x]], paste0(BiblioPath1, Output, "Stats/Statisitcs Table (Jul 23).xlsx"), sheetName = paste(FigNam[aa], Days[bb], x), append = TRUE))
  }
}
rm(aa, bb, Days)

gc()
```

